---
title: "Introduction to Single Cell RNAseq Part 1"
author: "UCD Bioinformatics Core"
output:
    html_document:
      keep_md: TRUE
---

Last Updated: July 10, 2022

# Part 1: Loading data from CellRanger into R

Our first Markdown document concentrates on getting data into R and setting up our initial object.

## Single Cell Analysis with Seurat and some custom code!

[Seurat](http://satijalab.org/seurat/) (now Version 4) is a popular R package that is designed for QC, analysis, and exploration of single cell data. Seurat aims to enable users to identify and interpret sources of heterogeneity from single cell transcriptomic measurements, and to integrate diverse types of single cell data. Further, the authors provide several [tutorials](https://satijalab.org/seurat/vignettes.html), on their website.

The **expression_data_cellranger.zip** file contains the single cell matrix files and HDF5 files for three single nuclei RNASeq samples from [Becker et al., 2022](https://www.nature.com/articles/s41588-022-01088-x)

We start each markdown document with loading needed libraries for R:


```{r, warning=FALSE,error=FALSE,message=FALSE}
# must have Seurat
library(Seurat)
library(kableExtra)
library(ggplot2)
```


### Setup the experiment folder and data info
```{r exp_setup, warning=FALSE,error=FALSE, message=FALSE}
experiment_name = "Colon Cancer"
dataset_loc <- "./expression_data_cellranger"
ids <- c("A001-C-007", "A001-C-104", "B001-A-301")
```


## Read in the cellranger sample metrics csv files

There is a single metrics summary file for each sample.

The code in the box below reads in a single metrics summary file. **To read in a summary file for each sample in your experiment, you should use the code in the next box instead.**

```{r, single_metrics}
experiment.metrics <- read.csv(file.path(dataset_loc, ids[1], "outs/metrics_summary.csv"))
sequencing.metrics <- data.frame(t(experiment.metrics[,c(1:19)]))
rownames(sequencing.metrics) <- gsub("\\.", " ", rownames(sequencing.metrics))
colnames(sequencing.metrics) <- "All samples"
sequencing.metrics %>%
  kable(caption = 'Cell Ranger Results') %>%
  pack_rows("Overview", 1, 3, label_row_css = "background-color: #666; color: #fff;") %>%
  pack_rows("Sequencing Characteristics", 4, 9, label_row_css = "background-color: #666; color: #fff;") %>%
  pack_rows("Mapping Characteristics", 10, 19, label_row_css = "background-color: #666; color: #fff;") %>%
  kable_styling("striped")
```

```{r read_metrics, eval=FALSE}
d10x.metrics <- lapply(ids, function(i){
  metrics <- read.csv(file.path(dataset_loc,paste0(i,"/outs"),"metrics_summary.csv"), colClasses = "character")
})
experiment.metrics <- do.call("rbind", d10x.metrics)
rownames(experiment.metrics) <- ids

sequencing.metrics <- data.frame(t(experiment.metrics[,c(1:19)]))

row.names(sequencing.metrics) <- gsub("\\."," ", rownames(sequencing.metrics))

sequencing.metrics %>%
  kable(caption = 'Cell Ranger Results') %>%
  pack_rows("Overview", 1, 3, label_row_css = "background-color: #666; color: #fff;") %>%
  pack_rows("Sequencing Characteristics", 4, 9, label_row_css = "background-color: #666; color: #fff;") %>%
  pack_rows("Mapping Characteristics", 10, 19, label_row_css = "background-color: #666; color: #fff;") %>%
  kable_styling("striped")
```


## Load the Cell Ranger Matrix Data and create the base Seurat object.
Cell Ranger provides a function `cellranger aggr` that will combine multiple samples into a single matrix file. However, when processing data in R this is unnecessary and we can quickly aggregate them in R.

Seurat provides a function `Read10X` and `Read10X_h5` to read in 10X data folder. First we read in data from each individual sample folder. 

Later, we initialize the Seurat object (`CreateSeuratObject`) with the raw (non-normalized data). Keep all cells with at least 200 detected genes. Also extracting sample names, calculating and adding in the metadata mitochondrial percentage of each cell. Adding in the metadata batchid and cell cycle. Finally, saving the raw Seurat object.

## Load the Cell Ranger Matrix Data (hdf5 file) and create the base Seurat object.
```{r load_data_hdf5, warning=FALSE,error=FALSE, message=FALSE}
d10x.data <- lapply(ids, function(i){
  d10x <- Read10X_h5(file.path(dataset_loc, i, "/outs","raw_feature_bc_matrix.h5"))
  colnames(d10x) <- paste(sapply(strsplit(colnames(d10x),split="-"),'[[',1L),i,sep="-")
  d10x
})
names(d10x.data) <- ids

str(d10x.data)
```

If you don't have the needed hdf5 libraries you can read in the matrix files like such

```{r load_data_matrix, eval=FALSE}
d10x.data <- sapply(ids, function(i){
  d10x <- Read10X(file.path(dataset_loc, i, "/outs","raw_feature_bc_matrix"))
  colnames(d10x) <- paste(sapply(strsplit(colnames(d10x), split="-"), '[[', 1L), i, sep="-")
  d10x
})
names(d10x.data) <- ids
```


Lets recreate the barcode rank plot from the Cell Ranger web summary file.

To color each barcode rank plot using the number of cells estimated by Cell Ranger, skip the code in the box below and use the next one instead.
```{r fig_barcode_umi, fig.width=10}
plot_cellranger_cells <- function(ind){
  xbreaks = c(1,1e1,1e2,1e3,1e4,1e5,1e6)
  xlabels = c("1","10","100","1000","10k","100K","1M")
  ybreaks = c(1,2,5,10,20,50,100,200,500,1000,2000,5000,10000,20000,50000,100000,200000,500000,1000000)
  ylabels = c("1","2","5","10","2","5","100","2","5","1000","2","5","10k","2","5","100K","2","5","1M")

  pl1 <- data.frame(index=seq.int(1,ncol(d10x.data[[ind]])),
                    nCount_RNA = sort(Matrix:::colSums(d10x.data[[ind]])+1,decreasing=T),
                    nFeature_RNA = sort(Matrix:::colSums(d10x.data[[ind]]>0)+1,decreasing=T)) %>%
    ggplot() + 
    scale_color_manual(values=c("red2","blue4"), labels=c("Features", "UMI"), name=NULL) +
    ggtitle(paste("CellRanger filltered cells:",ids[ind],sep=" ")) + xlab("Barcodes") + ylab("counts (UMI or Features") + 
    scale_x_continuous(trans = 'log2', breaks=xbreaks, labels = xlabels) + 
    scale_y_continuous(trans = 'log2', breaks=ybreaks, labels = ylabels) +
    geom_line(aes(x=index, y=nCount_RNA, color = "UMI"), size=1.75) +
    geom_line(aes(x=index, y=nFeature_RNA, color = "Features"), size=1.25)

  return(pl1)
}

plot_cellranger_cells(1)
plot_cellranger_cells(2)
plot_cellranger_cells(3)
```

```{r fig_barcode_umi_original, eval=FALSE}
cr_filtered_cells <- as.numeric(gsub(",","",as.character(unlist(sequencing.metrics["Estimated Number of Cells",]))))

plot_cellranger_cells <- function(ind){
  xbreaks = c(1,1e1,1e2,1e3,1e4,1e5,1e6)
  xlabels = c("1","10","100","1000","10k","100K","1M")
  ybreaks = c(1,2,5,10,20,50,100,200,500,1000,2000,5000,10000,20000,50000,100000,200000,500000,1000000)
  ylabels = c("1","2","5","10","2","5","100","2","5","1000","2","5","10k","2","5","100K","2","5","1M")

  pl1 <- data.frame(index=seq.int(1,ncol(d10x.data[[ind]])), nCount_RNA = sort(Matrix:::colSums(d10x.data[[ind]])+1,decreasing=T), nFeature_RNA = sort(Matrix:::colSums(d10x.data[[ind]]>0)+1,decreasing=T)) %>% ggplot() + 
    scale_color_manual(values=c("grey50","red2","blue4"), labels=c("UMI_Background", "Features", "UMI_Cells"), name=NULL) +
    ggtitle(paste("CellRanger filltered cells:",ids[ind],sep=" ")) + xlab("Barcodes") + ylab("counts (UMI or Features") + 
    scale_x_continuous(trans = 'log2', breaks=xbreaks, labels = xlabels) + 
    scale_y_continuous(trans = 'log2', breaks=ybreaks, labels = ylabels) +
    geom_line(aes(x=index, y=nCount_RNA, color=index<=cr_filtered_cells[ind] , group=1), size=1.75) +
    geom_line(aes(x=index, y=nFeature_RNA, color="Features", group=1), size=1.25)

  return(pl1)
}

plot_cellranger_cells(1)
plot_cellranger_cells(2)
plot_cellranger_cells(3)
```


### Create the Seurat object

Filter criteria: remove genes that do not occur in a minimum of 10 cells and remove cells that don't have a minimum of 200 features

```{r create_seurat_object, warning=FALSE,error=FALSE, message=FALSE}
experiment.data <- do.call("cbind", d10x.data)

experiment.aggregate <- CreateSeuratObject(
  experiment.data,
  project = experiment_name,
  min.cells = 10,
  min.features = 300,
  names.field = 2,
  names.delim = "\\-")

experiment.aggregate
str(experiment.aggregate)
```

### The percentage of reads that map to the mitochondrial genome

* Low-quality / dying cells often exhibit extensive mitochondrial contamination.
* We calculate mitochondrial QC metrics with the PercentageFeatureSet function, which calculates the percentage of counts originating from a set of features.
* We use the set of all genes, in mouse these genes can be identified as those that begin with 'mt', in human data they begin with MT.

```{r explore, warning=FALSE,error=FALSE, message=FALSE}
experiment.aggregate$percent.mito <- PercentageFeatureSet(experiment.aggregate, pattern = "^MT-")
summary(experiment.aggregate$percent.mito)
```

### Lets spend a little time getting to know the Seurat object.

The Seurat object is the center of each single cell analysis. It stores __all__ information associated with the dataset, including data, annotations, analyses, etc. The R function slotNames can be used to view the slot names within an object.

```{r explore2, warning=FALSE,error=FALSE, message=FALSE}
slotNames(experiment.aggregate)
```

```{r explore3, warning=FALSE,error=FALSE, message=FALSE}
head(experiment.aggregate[[]])
```

#### Question(s)

1. What slots are empty, what slots have data?
2. What columns are available in meta.data?
3. Look up the help documentation for subset?


## Finally, save the original object and view the object.

Original data set in Seurat class, with no filtering
```{r save, eval=TRUE}
save(experiment.aggregate,file="original_seurat_object.RData")
```

## Get the next Rmd file
```{r getnextfile, eval=FALSE}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2022-March-Single-Cell-RNA-Seq-Analysis/main/data_analysis/scRNA_Workshop-PART2.Rmd", "scRNA_Workshop-PART2.Rmd")
```

## Session Information
```{r sessioinfo}
sessionInfo()
```
