---
title: "Single Cell RNAseq Part 5"
author: "Bioinformatics Core"
output:
    html_document:
      keep_md: TRUE
---

# Part 5: Enrichment, Model-Based DE, and Cell-Type Identification

```{r setup_opt, warning=FALSE,error=FALSE,message=FALSE, echo = F}
options(width = 450)
```

## Load libraries
```{r load_library, warning=FALSE,error=FALSE,message=FALSE}

library(Seurat)
library(ggplot2)
library(limma)
library(topGO)
```

## Load the Seurat object
```{r load_rdata, warning=FALSE,error=FALSE,message=FALSE}
load("clusters_seurat_object.RData")
experiment.merged
Idents(experiment.merged) <- "finalcluster"
```

# 1. Gene Ontology (GO) Enrichment of Genes Expressed in a Cluster
[Gene Ontology](http://geneontology.org/docs/ontology-documentation/) provides a controlled vocabulary for describing gene products.  Here we use enrichment analysis to identify GO terms that are overrepresented among the gene expressed in cells in a given cluster. 

```{r GO, warning=FALSE,error=FALSE,message=FALSE}
cluster12 <- subset(experiment.merged, idents = '12')
expr <- as.matrix(GetAssayData(cluster12))
# Filter out genes that are 0 for every cell in this cluster
bad <- which(rowSums(expr) == 0)
expr <- expr[-bad,]

# Select genes that are expressed > 0 in at least half of cells
n.gt.0 <- apply(expr, 1, function(x)length(which(x > 0)))
expressed.genes <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.5)]
all.genes <- rownames(expr)

# define geneList as 1 if gene is in expressed.genes, 0 otherwise
geneList <- ifelse(all.genes %in% expressed.genes, 1, 0)
names(geneList) <- all.genes

# Create topGOdata object
	GOdata <- new("topGOdata",
		ontology = "BP", # use biological process ontology
		allGenes = geneList,
		geneSelectionFun = function(x)(x == 1),
              annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "symbol")
# Test for enrichment using Fisher's Exact Test
	resultFisher <- runTest(GOdata, algorithm = "elim", statistic = "fisher")
	GenTable(GOdata, Fisher = resultFisher, topNodes = 20, numChar = 60)
```
* Annotated: number of genes (out of all.genes) that are annotated with that GO term
* Significant: number of genes that are annotated with that GO term and meet our criteria for "expressed"
* Expected: Under random chance, number of genes that would be expected to be annotated with that GO term and meeting our criteria for "expressed"
* Fisher: (Raw) p-value from Fisher's Exact Test

# 2. Model-based DE analysis in limma
[limma](https://bioconductor.org/packages/release/bioc/html/limma.html) is an R package for differential expression analysis of bulk RNASeq and microarray data.  We apply it here to single cell data.

Limma can be used to fit any linear model to expression data and is useful for analyses that go beyond two-group comparisons.  A detailed tutorial of model specification in limma is available [here](https://ucdavis-bioinformatics-training.github.io/2021-June-RNA-Seq-Analysis/data_analysis/DE_Analysis_mm_with_quizzes) and in the [limma User's Guide](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf).

```{r}
# filter genes to those expressed in at least 10% of cells
keep <- rownames(expr)[which(n.gt.0/ncol(expr) >= 0.1)]
expr2 <- expr[keep,]

# Set up "design matrix" with statistical model
cluster12$proper.ident <- make.names(cluster12$orig.ident)
mm <- model.matrix(~0 + proper.ident + S.Score + G2M.Score + percent.mito + nFeature_RNA, data = cluster12[[]])
head(mm)
tail(mm)

# Fit model in limma
fit <- lmFit(expr2, mm)
head(coef(fit))

# Test 'B001-A-301' - 'A001-C-007'
contr <- makeContrasts(proper.identB001.A.301 - proper.identA001.C.007, levels = colnames(coef(fit)))
contr
fit2 <- contrasts.fit(fit, contrasts = contr)
fit2 <- eBayes(fit2)
out <- topTable(fit2, n = Inf, sort.by = "P")
head(out, 30)
```

### Output columns:
* logFC: log fold change (since we are working with Seurat's natural log transformed data, will be natural log fold change)
* AveExpr: Average expression across all cells in expr2
* t: logFC divided by its standard error
* P.Value: Raw p-value (based on t) from test that logFC differs from 0
* adj.P.Val: Benjamini-Hochberg false discovery rate adjusted p-value
* B: log-odds that gene is DE 



# BONUS: Cell type identification with scMRMA
[scMRMA]([https://academic.oup.com/nar/article/50/2/e7/6396893]) (single cell Multi-Resolution Marker-based Annotation Algorithm) classifies cells by iteratively clustering them then annotating based on a hierarchical external database.

The databases included with the current version are only for use with human and mouse, but a user-constructed hierarchichal database can be used. 

The package can be installed from [Github](https://github.com/JiaLiVUMC/scMRMA):

```{r}
# Remove hashes to run
# install.packages("devtools")
# devtools::install_github("JiaLiVUMC/scMRMA")
```

```{r, warning = FALSE}
suppressPackageStartupMessages(library(scMRMA))
result <- scMRMA(input = experiment.merged,
                 species = "Hs",
                 db = "panglaodb")
table(result$uniformR$annotationResult)

## Add cell types to metadata
experiment.merged <- AddMetaData(experiment.merged, result$uniformR$annotationResult, col.name = "CellType")
table(experiment.merged$CellType, experiment.merged$orig.ident)
table(experiment.merged$CellType, experiment.merged$finalcluster)
DimPlot(experiment.merged, group.by = "CellType", label = TRUE)
```

## Get the next Rmd file
```{r getnextfile, eval=FALSE}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2022-December-Single-Cell-RNA-Seq-Analysis/main/data_analysis/scRNA_Workshop-PART6.Rmd", "scRNA_Workshop-PART6.Rmd")
```

                 
## Session Information
```{r session_info, warning=FALSE,error=FALSE,message=FALSE}
sessionInfo()
```
